Learning Action Models from State Observations

### 2018-01-24 doc/ijcai-18/observations-learning.tex
new updates to the ijcai18 paper

\begin{abstract}
This paper presents a classical planning compilation for learning \strips\ action models from state observations. The compilation approach does not require observing the precise actions that produced the observations because such actions are determined by a planner. Furthermore, the presented compilation is extensible to assess how well a \strips\ action model matches a given set of observations, which enables {\em model recognition}. Last but not least, the paper evaluates the performance of the proposed approach by learning action models for a wide range of classical planning domains from the International Planning Competition and assessing the learned models with respect to (1) the corresponding reference models and (2), given observations test sets. 
\end{abstract}

@@

\begin{enumerate}
\item {\em Learning action models as planning}.  The practicality of the compilation approach allow us to report results over a wide range of planning domains. Furthermore, opens up a way towards the {\em bootstrapping} of planning action models (a planner exploring its state space and using the obtained data to learn/update its action model).
\item {\em Learning from state observations}.  Our approach does not require observing the precise actions that produced the observations because such actions are determined by a classical planner.
\item {\em Model evaluation}. The compilation is extensible to assess how well a \strips\ action model matches a given set of observations, which enables {\em model recognition}. 
\end{enumerate}

\section{Background}
Our approach for learning \strips\ action models is compiling the leaning task into a classical planning task with conditional effects.

@@

Let us also define $F_v$, a new set of fluents $F\cap F_v=\emptyset$, that results from instantiating $\Psi$ using only the objects in $\Omega_v$ and that defines the elements that can appear in an action schema. For the blocksworld, $F_v$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$), on($v_1,v_2$), on($v_2,v_1$), on($v_2,v_2$)\}}.

@@

Finally we define $F_v(\xi)\subseteq F_v$ as the subset of elements that can appear in a given action schema $\xi$. For instance, for the {\em stack} action schema $F_v({\tt stack})=F_v$ while $F_v({\tt pickup})$={\small\tt\{handempty, holding($v_1$), clear($v_1$), ontable($v_1$), on($v_1,v_1$)\}} only contains the fluents from $F_v$ that do not involve $v_2$ because the action header contains the single parameter $v_1$.

\begin{figure}
\begin{footnotesize}
\begin{verbatim}
(:action stack
 :parameters (?v1 ?v2 - object)
 :precondition (and (holding ?v1) (clear ?v2))
 :effect (and (not (holding ?v1)) 
              (not (clear ?v2))
              (handempty) (clear ?v1) 
              (on ?v1 ?v2)))
\end{verbatim}
\end{footnotesize}
 \caption{\small \strips\ operator schema coding, in PDDL, the {\em stack} action from the {\em blocksworld}.}
\label{fig:stack}
\end{figure}

\section{Learning \strips\ action models}
Learning \strips\ action models from fully available input knowledge, i.e. from plans where every action in the plan is available as well as its corresponding {\em pre-} and {\em post-states}, is straightforward. In this case, \strips\ operator schemes are derived lifting the literals that change between the pre and post-state of the corresponding action executions. Preconditions are derived lifting the minimal set of literals that appears in all the pre-states of the corresponding actions.

In this paper we address a more challenging learning task, where less input knowledge is available. The addressed learning task corresponds to observing an agent acting in the world but watching only the results of its plan executions, the actual executed actions are unobserved. This learning task is defined as $\Lambda=\tup{\Psi,\Xi,\mathcal{O}}$, where:
\begin{itemize}
\item $\Psi$ is the set of predicates that define the abstract state space of a given classical planning frame.
\item $\Xi$ is the set of empty operator schemes, wherein each $\xi\in\Xi$ is only composed of $head(\xi)$.
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is a sequence of {\em state observations} obtained observing the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$.
\end{itemize}

A solution to $\Lambda$ is a set of operator schema $\Xi'$ compliant with the predicates in $\Psi$, the headers in $\Xi$ and the given sequence of state observations $\mathcal{O}$. A planning compilation is a suitable approach for addressing a $\Lambda$ learning task because a solution must not only determine the \strips\ action model $\Xi'$ but also, the {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$, that explains $\mathcal{O}$. Figure~\ref{fig:lexample} shows an example of a $\Lambda$ task for learning a \strips\ action model in the blocksworld from the sequence of five state observations that corresponds to inverting a 2-blocks tower.

\begin{figure}
{\footnotesize\tt ;;; Predicates in $\Psi$}
\begin{footnotesize}
\begin{verbatim}
(handempty) (holding ?o  - object)
(clear ?o - object) (ontable ?o - object)
(on ?o1 - object ?o2 - object)
\end{verbatim}
\end{footnotesize}
\vspace{0.2cm}
{\footnotesize\tt ;;; Headers in $\Xi$}
\begin{footnotesize}
\begin{verbatim}
(pickup v1) (putdown v1)
(stack v1 v2} (unstack v1 v2)
\end{verbatim}
\end{footnotesize}
\vspace{0.2cm}
{\footnotesize\tt ;;; Observations in $\mathcal{O}$}
\begin{footnotesize}
\begin{verbatim}
;;; observation #0
(clear block2) (on block2 block1)
(ontable block1) (handempty)

@@

Our approach for addressing the learning task is compiling $\Lambda$ into a classical planning task with conditional effects. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:

\begin{enumerate}
\item Programs the \strips\ action model $\Xi'$. A solution plan starts with a {\em prefix} that, for each $\xi\in\Xi$, determines which fluents $f\in F_v(\xi)$ belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item Validates the \strips\ action model $\Xi'$ in $\mathcal{O}$. The solution plan continues with a postfix that produces the given sequence of states $\tup{s_0,s_1,\ldots,s_{n}}$ using the programmed action model $\Xi'$.
\end{enumerate}

Given a learning task $\Lambda=\tup{\Psi,\Xi,\mathcal{O}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents $F$ that is built instantiating the predicates $\Psi$ with the objects appearing in the input observations $\mathcal{O}$.
\item Fluents representing the programmed action model $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative effect/positive effect in the \strips\ operator schema $\xi\in \Xi$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) are represented by the fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to {\em True}.
\item Fluent $mode_{prog}$ indicating whether the operator schemes are programmed or validated (already programmed) and fluents $\{test_i\}_{1\leq i\leq n}$, indicating the observation where the action model is validated.
\end{itemize}
\item $I_{\Lambda}$ contains the fluents from $F$ that encode $s_0$ (the first observation) and $mode_{prog}$ set to true. In addition, our compilation assumes that initially operator schemes are programmed with every possible precondition, no negative effect and no positive effect. With this regard, the fluents $pre_f(\xi)$ hold at the initial state for every $f\in F_v(\xi)$.
\item $G_{\Lambda}=\bigcup_{1\leq i\leq n}\{test_i\}$, indicates that the programmed action model is validated in all the input observations.
\item $A_{\Lambda}$ comprises three kinds of actions:
\begin{enumerate}
\item Actions for {\em programming} operator schema $\xi\in\Xi$:
\begin{itemize}
\item Actions for {\bf removing} a {\em precondition} $f\in F_v(\xi)$ from the action schema $\xi\in\Xi$.

@@

\item Actions for {\em applying} an already programmed operator schema $\xi\in\Xi$ bound with the objects $\omega\subseteq\Omega^{ar(\xi)}$. We assume that the operators headers are known so the binding of the operator schema is done implicitly by order of appearance of the action parameters, i.e. variables $pars(\xi)$ are bound to the objects in $\omega$ appearing at the same position. Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying a programmed operator $stack$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\}.
\end{align*}
\end{small}

@@

\begin{figure}[hbt!]
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_on_stack_v1_v1)) (on ?o1 ?o1))
        (or (not (pre_on_stack_v1_v2)) (on ?o1 ?o2))
        (or (not (pre_on_stack_v2_v1)) (on ?o2 ?o1))
        (or (not (pre_on_stack_v2_v2)) (on ?o2 ?o2))
        (or (not (pre_ontable_stack_v1)) (ontable ?o1))
        (or (not (pre_ontable_stack_v2)) (ontable ?o2))
        (or (not (pre_clear_stack_v1)) (clear ?o1))
        (or (not (pre_clear_stack_v2)) (clear ?o2))
        (or (not (pre_holding_stack_v1)) (holding ?o1))
        (or (not (pre_holding_stack_v2)) (holding ?o2))
        (or (not (pre_handempty_stack)) (handempty)))
  :effect
   (and (when (del_on_stack_v1_v1) (not (on ?o1 ?o1)))
        (when (del_on_stack_v1_v2) (not (on ?o1 ?o2)))
        (when (del_on_stack_v2_v1) (not (on ?o2 ?o1)))
        (when (del_on_stack_v2_v2) (not (on ?o2 ?o2)))
        (when (del_ontable_stack_v1) (not (ontable ?o1)))
        (when (del_ontable_stack_v2) (not (ontable ?o2)))
        (when (del_clear_stack_v1) (not (clear ?o1)))
        (when (del_clear_stack_v2) (not (clear ?o2)))
        (when (del_holding_stack_v1) (not (holding ?o1)))
        (when (del_holding_stack_v2) (not (holding ?o2)))
        (when (del_handempty_stack) (not (handempty)))
        (when (add_on_stack_v1_v1) (on ?o1 ?o1))
        (when (add_on_stack_v1_v2) (on ?o1 ?o2))
        (when (add_on_stack_v2_v1) (on ?o2 ?o1))
        (when (add_on_stack_v2_v2) (on ?o2 ?o2))
        (when (add_ontable_stack_v1) (ontable ?o1))
        (when (add_ontable_stack_v2) (ontable ?o2))
        (when (add_clear_stack_v1) (clear ?o1))
        (when (add_clear_stack_v2) (clear ?o2))
        (when (add_holding_stack_v1) (holding ?o1))
        (when (add_holding_stack_v2) (holding ?o2))
        (when (add_handempty_stack) (handempty))
        (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small Action for applying an already programmed schema $stack$ as encoded in PDDL (implications coded as disjunctions).}
\label{fig:compilation}
\end{figure}

@@

\begin{lemma}
Soundness. Any classical plan $\pi$ that solves $P_{\Lambda}$ induces an action model $\Xi'$ that solves the learning task $\Lambda$.
\end{lemma}

\begin{proof}[Proof sketch]
\begin{small}
The compilation forces that once the preconditions of an operator schema $\xi \in \Xi'$ are programmed, they cannot be altered. The same happens with the positive and negative effects. Furthermore because of the preconditions of the $\mathsf{programPre_{f,\xi}}$ actions, effects are only programmable after preconditions are programmed. Once the operator schemes $\Xi'$ are programmed, they can only be applied because of the $mode_{prog}$ fluent. To solve $P_{\Lambda}$, the goals $\{test_i\}$, {\small $1\leq i\leq n$} can only be achieved: executing an applicable sequence of programmed operator schemes that reaches every state $s_i\in\mathcal{O}$, starting from $s_0$ and following  the sequence {\small $1\leq i\leq n$}. Therefore if $test_n$ is achieved, it means that the programmed action model $\Xi'$ is compliant with the provided input knowledge and hence, solves $\Lambda$.
\end{small}
\end{proof}

@@

\begin{proof}[Proof sketch]
\begin{small}
By definition, $F_v(\xi)\subseteq F_\Lambda$ fully captures the set of elements that can appear in a \strips\ action schema $\xi\in\Xi$. In addition, any possible \strips\ action schema $\Xi'$ that can be built with the fluents in $F_v$ can be computed  with the $P_{\Lambda}$ compilation. The only \strips\ action models that the compilation cannot compute are the ones that violate the state constraints defined by the $\mathcal{O}$ sequence, which constrain as well the solutions to the $\Lambda$ learning task.
\end{small}
\end{proof}

\section{Evaluating \strips\ action models}
The compilation can be extended to evaluate how well a \strips\ action model matches a given set of observations. This allows to asses the quality of learned models without a reference model and enables the recognition of \strips\ action models.

The main idea behind the compilation extension is that how well a \strips\ action model $\Xi$ explains a given sequence of observations $\mathcal{O}$, depends on the amount of {\em edition} that one has to introduce into $\Xi$ to produce $\mathcal{O}$ (in the extreme if $\Xi$ perfectly explains $\mathcal{O}$, no {\em edit} must be done).

\subsection{The edit distance}
Let us define first the \emph{operations} allowed to edit a given \strips\ action model. With the aim of keeping tractable the branching factor of the planning instance that results from our compilation, we only define two {\em edit operations}:
\begin{itemize}
\item {\em Deletion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$, such that $f\in F_v(\xi)$, is removed from the operator schema $\xi\in\Xi$.
\item {\em Insertion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$, such that $f\in F_v(\xi)$, is added to the operator schema $\xi\in\Xi$.
\end{itemize}

@@

Since $F_v$ is a bound set, the maximum number of edits that can be introduced to a given action model defined within $F_v$ is bound as well. In more detail, for an operator schema $\xi\in\Xi$ the maximum number of edits that can be introduced to their precondition set is $|F_v(\xi)|$. With regard to the effects, the maximum number of edits that can be introduced are two times $|F_v(\xi)|$.

\begin{definition}
Let $\Xi$ be an action model built from the set of possible elements $F_v$. The \textbf{maximum edit distance}, defined as $\delta(\Xi,*)=\sum_{\xi\in\Xi} 3|F_v(\xi)|$, is an upper bound on the distance from $\Xi$ to any \strips\ action model definable within $F_v$.
\end{definition}

Finally, we are ready to define the distance of a model $\Xi$ to an observation sequence $\mathcal{O}$ on the basis of the \emph{edit distance} between two models.

\begin{definition}
Given an action model $\Xi$ built from the set of possible elements $F_v$ and a sequence of observations $\mathcal{O}$. The {\bf observation edit distance}, denoted by  $\delta(\Xi,\mathcal{O})$, is the minimal edit distance to transform $\Xi$ into a model $\Xi'$ which: (1) is definable within $F_v$ and (2), can produce a plan $\pi=\tup{a_1, \ldots, a_n}$ such that $\pi$ induces the observation sequence $\mathcal{O}=\tup{s_0, s_1, \ldots, s_n}$; i.e., \[\delta(\Xi,\mathcal{O})=\min_{\exists\ \Xi' s.t. \Xi' \rightarrow \mathcal{O}} \delta(\Xi,\Xi')\]
\end{definition}

The {\em observation edit distance} asses the quality of a learned strips action model with respect to a sequence of observations (that acts as a test set). The lower this distance, the better the learned model. The semantic nature of this evaluation is robust to learning episodes where actions are reformulated and still compliant with the inputs (e.g. the {\em blocksworld} operator {\small\tt stack} could be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa or the roles of actions parameters with the same type could be interchanged).

Given a set of possible \strips\ models and a sequence of state observations, the {\em recognition of \strips\ models} is the task of computing the model with the highest probability according to the cited observations. The {\em observations edit distance} is also helpful to compute the probability distribution of the possible \strips\ models given a sequence of state observations. The main idea, taken from {\em plan recognition as planning}~\cite{ramirez2009plan}, is to map distances into likelihoods and use the Bayes rule.

According to the Bayes rule, the probability of an hypothesis $\mathcal{H}$ given the observations $\mathcal{O}$ can be computed with $P(\mathcal{H}|\mathcal{O})=\frac{P(\mathcal{O}|\mathcal{H})P(\mathcal{H})}{P(\mathcal{O})}$. In our scenario, the hypotheses are about the possible \strips\ action models that can be built within a given set of predicates $\Psi$ and a given a set of operator headers (in other words, given the $F_v(\xi)$ sets). Moreover, we assume that $P(\mathcal{O}|\mathcal{H})$ is given by the {\em observation edit distance}, mapping distances into probabilities, according to the following expression $P(\mathcal{O}|\Xi)=1-\frac{\delta(\Xi,\mathcal{O})}{\delta(\Xi,*)}$. The larger the distance the lower the likelihood.

>>>>

\begin{abstract}
This paper presents a classical planning compilation for learning \strips\ action models from state observations. The compilation approach does not require observing the precise actions that produced the observations because such actions are determined by a planner. Furthermore, the presented compilation is extensible to assess how well a \strips\ action model matches a given set of observations. Last but not least, the paper evaluates the performance of the proposed approach by learning action models for a wide range of classical planning domains from the International Planning Competition and assessing the learned models with respect to (1) the corresponding reference models and (2), given observations test sets. 
\end{abstract}

@@

\begin{enumerate}
\item {\em Learning action models as planning}.  The practicality of the compilation approach allow us to report results over a wide range of planning domains. Furthermore, opens up a way towards the {\em bootstrapping} of planning action models (a planner exploring its state space and using the obtained data to learn/update its action model).
\item {\em Learning from state observations}.  Our approach does not require the learning samples to be observations of actions but simply state observations, broadening the range of application to external observers.
\item {\em Model evaluation}. The compilation is extensible to assess how well a \strips\ action model matches a given set of observations, which enables {\em model recognition}. 
\end{enumerate}

\section{Background}
Our approach for learning \strips\ action models is compiling this leaning task into a classical planning task with conditional effects.

@@

Let us define $F_v$, a new set of fluents $F\cap F_v=\emptyset$, that results from instantiating $\Psi$ using only the objects in $\Omega_v$ and that defines the elements that can appear in an action schema. For the blocksworld, $F_v$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$), on($v_1,v_2$), on($v_2,v_1$), on($v_2,v_2$)\}}.

@@

Finally, we define $F_v(\xi)\subseteq F_v$ as the subset of elements that can appear in a given action schema $\xi$ and that confine the space of possible action models. For instance, for the {\em stack} action schema $F_v({\tt stack})=F_v$ while $F_v({\tt pickup})$={\small\tt\{handempty, holding($v_1$), clear($v_1$), ontable($v_1$), on($v_1,v_1$)\}} only contains the fluents from $F_v$ that do not involve $v_2$ because the action header contains the single parameter $v_1$.

\begin{figure}
\begin{footnotesize}
\begin{verbatim}
(:action stack
 :parameters (?v1 ?v2 - object)
 :precondition (and (holding ?v1) (clear ?v2))
 :effect (and (not (holding ?v1)) 
              (not (clear ?v2))
              (handempty) (clear ?v1) 
              (on ?v1 ?v2)))
\end{verbatim}
\end{footnotesize}
 \caption{\small \strips\ operator schema coding, in PDDL, the {\em stack} action from a four-operator {\em blocksworld}.}
\label{fig:stack}
\end{figure}

\section{Learning \strips\ action models}
Learning \strips\ action models from plans where every action in the plan is available as well as its corresponding {\em pre-} and {\em post-states}, is straightforward. In this case, \strips\ operator schemes are derived lifting the literals that change between the pre and post-state of the corresponding action executions. Preconditions are derived lifting the minimal set of literals that appears in all the corresponding pre-states.

This paper addresses a more challenging learning task, where less input knowledge is available. The addressed learning task corresponds to observing an agent acting in the world but watching only the results of its plan executions, the actual executed actions are unobserved. This learning task is defined as $\Lambda=\tup{\Xi,\Psi,\mathcal{O}}$, where:
\begin{itemize}
\item $\Xi$ is the set of empty operator schemes, wherein each $\xi\in\Xi$ is only composed of $head(\xi)$.
\item $\Psi$ is the set of predicates, that define the abstract state space of a given classical planning frame.
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is a sequence of {\em state observations} obtained observing the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$.
\end{itemize}

A solution to $\Lambda$ is a set of operator schema $\Xi'$ compliant with the headers $\Xi$, the predicates $\Psi$, and the given sequence of state observations $\mathcal{O}$. A planning compilation is a suitable approach for addressing a $\Lambda$ learning task because a solution must not only determine the \strips\ action model $\Xi'$ but also, the {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$, that explains $\mathcal{O}$. Figure~\ref{fig:lexample} shows a $\Lambda$ task for learning a \strips\ action model in the blocksworld from the sequence of five state observations that corresponds to inverting a 2-blocks tower.

\begin{figure}
{\footnotesize\tt ;;;;;; Headers in $\Xi$}
\begin{footnotesize}
\begin{verbatim}
(pickup v1) (putdown v1)
(stack v1 v2} (unstack v1 v2)
\end{verbatim}
\end{footnotesize}
\vspace{0.2cm}
{\footnotesize\tt ;;;;;; Predicates $\Psi$}
\begin{footnotesize}
\begin{verbatim}
(handempty) (holding ?o  - object)
(clear ?o - object) (ontable ?o - object)
(on ?o1 - object ?o2 - object)
\end{verbatim}
\end{footnotesize}
\vspace{0.2cm}
{\footnotesize\tt ;;;;;; Observations $\mathcal{O}$}
\begin{footnotesize}
\begin{verbatim}
;;; observation #0
(clear block2) (on block2 block1)
(ontable block1) (handempty)

@@

Our approach for addressing the learning task is compiling $\Lambda$ into a classical planning task $P_{\Lambda}$ with conditional effects. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:

\begin{enumerate}
\item {\bf Programs the \strips\ action model $\Xi'$}. A solution plan starts with a {\em prefix} that, for each $\xi\in\Xi$, determines which fluents $f\in F_v(\xi)$ belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item {\bf Validates the \strips\ action model $\Xi'$ in $\mathcal{O}$}. The solution plan continues with a postfix that produces the given sequence of states $\tup{s_0,s_1,\ldots,s_{n}}$ using the programmed action model $\Xi'$.
\end{enumerate}

Given a learning task $\Lambda=\tup{\Xi,\Psi,\mathcal{O}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents $F$ built instantiating the predicates $\Psi$ with the objects appearing in the input observations $\mathcal{O}$.
\item Fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$, that represent the programmed action model. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative effect/positive effect in the \strips\ operator schema $\xi\in \Xi'$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) is represented by the pair of fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to {\em True}.
\item Fluent $mode_{prog}$ indicating whether the operator schemes are programmed or validated (already programmed) and fluents $\{test_i\}_{1\leq i\leq n}$, indicating the observation where the action model is validated.
\end{itemize}
\item $I_{\Lambda}$ contains the fluents from $F$ that encode $s_0$ (the first observation) and $mode_{prog}$ set to true. Our compilation assumes that initially operator schemes are programmed with every possible precondition, no negative effect and no positive effect. Therefore fluents $pre_f(\xi)$, for every $f\in F_v(\xi)$, hold also at the initial state.
\item $G_{\Lambda}=\bigcup_{1\leq i\leq n}\{test_i\}$, indicates that the programmed action model is validated in all the input observations.
\item $A_{\Lambda}$ comprises three kinds of actions:
\begin{enumerate}
\item Actions for {\em programming} operator schema $\xi\in\Xi$:
\begin{itemize}
\item Actions for {\bf removing} a {\em precondition} $f\in F_v(\xi)$ from the action schema $\xi\in\Xi$.

@@

\item Actions for {\em applying} an already programmed operator schema $\xi\in\Xi$ bound with the objects $\omega\subseteq\Omega^{ar(\xi)}$. Given that the operators headers are known, the variables $pars(\xi)$ are bound to the objects in $\omega$ appearing at the same position. Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying a programmed operator $stack$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\}.
\end{align*}
\end{small}

@@

\begin{figure}[hbt!]
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_on_stack_v1_v1)) (on ?o1 ?o1))
        (or (not (pre_on_stack_v1_v2)) (on ?o1 ?o2))
        (or (not (pre_on_stack_v2_v1)) (on ?o2 ?o1))
        (or (not (pre_on_stack_v2_v2)) (on ?o2 ?o2))
        (or (not (pre_ontable_stack_v1)) (ontable ?o1))
        (or (not (pre_ontable_stack_v2)) (ontable ?o2))
        (or (not (pre_clear_stack_v1)) (clear ?o1))
        (or (not (pre_clear_stack_v2)) (clear ?o2))
        (or (not (pre_holding_stack_v1)) (holding ?o1))
        (or (not (pre_holding_stack_v2)) (holding ?o2))
        (or (not (pre_handempty_stack)) (handempty)))
  :effect
   (and (when (del_on_stack_v1_v1) (not (on ?o1 ?o1)))
        (when (del_on_stack_v1_v2) (not (on ?o1 ?o2)))
        (when (del_on_stack_v2_v1) (not (on ?o2 ?o1)))
        (when (del_on_stack_v2_v2) (not (on ?o2 ?o2)))
        (when (del_ontable_stack_v1) (not (ontable ?o1)))
        (when (del_ontable_stack_v2) (not (ontable ?o2)))
        (when (del_clear_stack_v1) (not (clear ?o1)))
        (when (del_clear_stack_v2) (not (clear ?o2)))
        (when (del_holding_stack_v1) (not (holding ?o1)))
        (when (del_holding_stack_v2) (not (holding ?o2)))
        (when (del_handempty_stack) (not (handempty)))
        (when (add_on_stack_v1_v1) (on ?o1 ?o1))
        (when (add_on_stack_v1_v2) (on ?o1 ?o2))
        (when (add_on_stack_v2_v1) (on ?o2 ?o1))
        (when (add_on_stack_v2_v2) (on ?o2 ?o2))
        (when (add_ontable_stack_v1) (ontable ?o1))
        (when (add_ontable_stack_v2) (ontable ?o2))
        (when (add_clear_stack_v1) (clear ?o1))
        (when (add_clear_stack_v2) (clear ?o2))
        (when (add_holding_stack_v1) (holding ?o1))
        (when (add_holding_stack_v2) (holding ?o2))
        (when (add_handempty_stack) (handempty))
        (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small Action for applying an already programmed schema $stack$ as encoded in PDDL (implications are coded as disjunctions).}
\label{fig:compilation}
\end{figure}

@@

\begin{lemma}
Soundness. Any classical plan $\pi$ that solves $P_{\Lambda}$ induces an action model $\Xi'$ that solves the $\Lambda$ learning task.
\end{lemma}

\begin{proof}[Proof sketch]
\begin{small}
The compilation forces that once the preconditions of an operator schema $\xi \in \Xi'$ are programmed, they cannot be altered. The same happens with the positive and negative effects. Furthermore, once operator schemes $\Xi'$ are programmed, they can only be applied because of the $mode_{prog}$ fluent. Finally, $P_{\Lambda}$ is only solvable if fluents $\{test_i\}$, {\small $1\leq i\leq n$} hold at the last reached state. These goals can only be achieved executing an applicable sequence of programmed operator schemes that reaches every state $s_i\in\mathcal{O}$, starting from $s_0$ and following the sequence {\small $1\leq i\leq n$} which means that the programmed action model $\Xi'$ complies with the provided observations and hence, solves $\Lambda$.
\end{small}
\end{proof}

@@

\begin{proof}[Proof sketch]
\begin{small}
The compilation does not discard any possible \strips\ action schema definable within $F_v$ that satisfy the state trajectory constraints given by the $\mathcal{O}$ sequence. By definition, $F_v(\xi)\subseteq F_\Lambda$ fully captures the full set of elements that can appear in a \strips\ action schema $\xi\in\Xi$ given its header and the set of predicates $\Psi$. 
\end{small}
\end{proof}

The size of the compiled classical planning instances depends on the number of input examples. The larger the number of examples, the larger the compilation (a larger number of "test" fluents and "validate" actions), and also the more actions the solution plan requires to derive the action model.

\section{Evaluation of \strips\ action models}
How well a \strips\ action model $\Xi$ explains a given sequence of observations $\mathcal{O}$ can be assessed according to the amount of {\em edition} required by $\Xi$ to produce $\mathcal{O}$ (in the extreme, if $\Xi$ perfectly explains $\mathcal{O}$, then no {\em edition} of the model is necessary).

\subsection{Edition of \strips\ action models}
Let us define first the \emph{operations} allowed to edit a given \strips\ action model. With the aim of keeping tractable the branching factor of the planning instance that results from our compilation, we only define two {\em edit operations}:
\begin{itemize}
\item {\em Deletion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$, such that $f\in F_v(\xi)$, is removed from the operator schema $\xi\in\Xi$.
\item {\em Insertion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$, such that $f\in F_v(\xi)$, is added to the operator schema $\xi\in\Xi$.
\end{itemize}

@@

Since $F_v$ is a bound set, the maximum number of edits that can be introduced to a given action model defined within $F_v$ is bound as well. In more detail, for an operator schema $\xi\in\Xi$ the maximum number of edits that can be introduced to their precondition set is $|F_v(\xi)|$ while the max number of edits that can be introduced to the effects are two times $|F_v(\xi)|$.

\begin{definition}
Let $\Xi$ be an action model built from the set of possible elements $F_v$. The \textbf{maximum edit distance} is $\delta(\Xi,*)=\sum_{\xi\in\Xi} 3|F_v(\xi)|$ and sets an upperbound on the edit distance from $\Xi$ to any other \strips\ action model definable within $F_v$.
\end{definition}

\subsection{The observation edit distance}
We define now an edit distance to asses the quality of learned models with respect to a sequence of observations (that acts as a test set).

\begin{definition}
Given an action model $\Xi$ built from the set of possible elements $F_v$ and a sequence of observations $\mathcal{O}$. The {\bf observation edit distance}, denoted by  $\delta(\Xi,\mathcal{O})$, is the minimal edit distance from $\Xi$ to a model $\Xi'$ such that: (1) $\Xi'$ is also definable within $F_v$ and (2), $\Xi'$ can produce a plan $\pi=\tup{a_1, \ldots, a_n}$ that induces $\mathcal{O}=\tup{s_0, s_1, \ldots, s_n}$, the given observations sequence; i.e., \[\delta(\Xi,\mathcal{O})=\min_{\forall \Xi' \rightarrow \mathcal{O}} \delta(\Xi,\Xi')\]
\end{definition}

Because its semantic nature, the {\em observation edit distance} is robust to learning episodes where actions are reformulated and still compliant with the inputs (e.g. the {\em blocksworld} operator {\small\tt stack} could be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa or the roles of the action parameters could be interchanged).

Furthermore the {\em observation edit distance} enables the recognition of \strips\ action models. The idea, taken from {\em plan recognition as planning}~\cite{ramirez2009plan}, is to map distances into likelihoods. We can map the {\em observation edit distance} into a likelihood with the following expression $P(\mathcal{O}|\Xi)=1-\frac{\delta(\Xi,\mathcal{O})}{\delta(\Xi,*)}$. The larger the {\em observation edit distance} the lower this likelihood.

%Given a set of possible \strips\ models and a sequence of state observations, the {\em recognition of \strips\ models} is the task of computing the model with the highest probability according to the cited observations. According to the Bayes rule, the probability of an hypothesis $\mathcal{H}$ given the observations $\mathcal{O}$ can be computed with $P(\mathcal{H}|\mathcal{O})=\frac{P(\mathcal{O}|\mathcal{H})P(\mathcal{H})}{P(\mathcal{O})}$. In our scenario, the hypotheses are about the possible \strips\ action models that can be built within a given set of predicates $\Psi$ and a given a set of operator headers (in other words, given the $F_v(\xi)$ sets). With this regard, $P(\Xi|\mathcal{O})$, the probability distribution of the possible \strips\ models (within the $F_v(\xi)$ sets) given an observation sequence $\mathcal{O}$ could be computed by:
%\begin{enumerate}
%\item Computing the {\em observation edit distance} $\delta(\Xi,\mathcal{O})$ for every possible model $\Xi$.
%\item Applying the resulting distances to the above $P(\mathcal{O}|\Xi)$ formula to map these distances into likelihoods
%\item Applying the Bayes rule to obtain the normalized posterior probabilities, these probabilities must sum 1.
%\end{enumerate}
